---
title: "How to keep it togetheR"
author: "Marc A.T. Teunis, Ph.D. - Hogeschool Utrecht, Lab. Innovative Testing"
date: "Last update: `r Sys.time()`"
output:
  pdf_document: default
  html_document: default
  ioslides_presentation:
    highlight: pygments
    widescreen: yes
  beamer_presentation:
    highlight: pygments
subtitle: "_List-columns in a dataframe_"
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE, 
                      error = FALSE, 
                      fig.width = 4,
                      fig.height = 4)
```

```{r, root, include=FALSE}
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)

```

# Demo for the Utrecht University R Cafe, 28 January 2019

## What is R to me?

 - Every day go-to for analytics of very different types of data
 - Statistical analysis `{lme}`, `{lme4}`, `{nlme}`
 - Genomics `{DESeq2}`, `{edgeR}`, `{limma}`, `{Glimma}`
 - Microbiome analysis `Bioconductor` & `qiime2`
 - Interactivity with `{Shiny}` and `{flexdashboards}`
 - Reproducibility `{rmarkdown}`, `{bookdown}`, `{blogdown}`
 - Tutorials and teaching `{learnr}`, `{reticulate}` -- combine Python & R
 - Getting data `{plumber}`, `{getGEO}`, `{rentrez}`
 - Text mining `{tidytext}`, `{igraph}`, `{ggiraph}`
 - Visualizations `{ggplot2}`, `{tidygraph}`, `{gganimate}`, `{tmap}`

## Content for today

 PART I: Starting with List-columns 
 
**There is more in this demo than we can cover today**

 PART II: MORE LIST-COLUMNS (do-it-yourself) 

## 
```{r, echo=FALSE, out.width= "1000px"}
knitr::include_graphics(path = file.path(root,
                                         "images",
                                         "ground_to_cover.jpg"))
```

## Getting access and materials

Clone the repository to your RStudio Environment from:

https://github.com/uashogeschoolutrecht/rcafe or login

http://rserverkcgdl.hudatascience.nl with 

login: ...
passwd: ...

## Packages
The packages used in this tutorial
```{r, packages}
library(tidyverse)
library(modelr)
library(lubridate)
library(broom)
library(purrr)
library(repurrrsive)
```

## Prequisites
See "resources.html" for tips on learning:

 - Import data with `{readr}` 
 - Tidy data with `{tidyr}` 
 - Data wrangling with  `{dplyr}` 
 - Using `lapply()` and/or `map()` for loops
 - Visualize data with `{ggplot2}` 

## Dataframes and lists are recursive vectors
```{r, recursive_vectors}
table1 <- tribble(
  ~a,    ~b,  ~c,     ~d, 
  "x",   1,   TRUE,   1.45, 
  "y",   2,   FALSE,  3.88,
  "z",   3,   TRUE,   33.5
  ) 
table1
is.atomic(table1$a)
```

## Column containing a list (in a dataframe)
```{r}
table2 <- tribble(
  ~ a,      ~b,      ~c,       ~d,     ~e, 
  "x",      1,       TRUE,     1.45,   1:10,
  "y",      2,       FALSE,    3.88,   c(TRUE, FALSE),
  "z",      3,       TRUE,     33.5,   "Utrecht"   
  ) 

is.list(table2$e)
is.vector(table2$e)
```

## Iterate over a dataframe
```{r}
map(table1, class)
```

## Iterate over a list-column
```{r}
map(table2$e, nchar)
```

## Case data

Let's switch to RStudio and open the file: 

`demo.Rmd`

## Data origin

Whooping cough outbreaks from The World Health Organization

http://data.euro.who.int/cisid/?TabID=463987

See for variable and data collection also: 

http://ecdc.europa.eu/sites/portal/files/documents/Pertussis%20AER.pdf

for more details see file: "load_data.R"

## Load the tidy version of the dataset
The code for cleaning and tidying the data is in the file "./load_data.R"
```{r, read_data}
source(file = file.path(root,
                        "load_data.R"))
  
head(pertussis_data_tidy, n = 2)
names(pertussis_data_tidy)
```

## Overall trend
```{r, echo=FALSE, fig.width=6, fig.height=4}
pertussis_data_tidy %>%
  group_by(year) %>%
  summarise(mean_annual_cases = mean(annual_pertussis_cases,
                                     na.rm = TRUE)) %>%
  ggplot(aes(x = year,
             y = mean_annual_cases)) + 
  geom_point() +
  geom_line() +
  ggtitle("Overall mean annual Whooping Cough cases per year")
```

## Data for individual countries, over time
```{r, echo=FALSE, fig.width=6, fig.height=4}
pertussis_data_tidy %>%
  na.omit() %>%
  group_by(country) %>%
  ggplot(aes(x = year,
             y = log10(annual_pertussis_cases))) +
  geom_line(aes(group = country, colour = country))  +
  theme_bw() +
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Data for only The Netherlands
```{r, echo=FALSE, fig.width=6, fig.height=4}
## only Netherlands data
netherlands_data <- pertussis_data_tidy %>%
  dplyr::filter(country == "Netherlands") %>%
  na.omit() ## for sake of simplicity we remove the years that are missing

netherlands_data %>%
  ggplot(aes(x = year,
             y = annual_pertussis_cases)) + 
  geom_line() +
  geom_point(colour = "red") +
  ggtitle("The Netherlands; Whooping Cough cases")
```

## Plot linear model for NL
```{r, echo=FALSE, fig.width=6, fig.height=4}
netherlands_model <- lm(
  annual_pertussis_cases ~ year, 
  data = netherlands_data %>%
    na.omit()
)
## plot model for NL

netherlands_data %>%
  na.omit() %>%
  add_predictions(netherlands_model) %>%
  ggplot(aes(x = year, 
             y = pred)) +
  geom_line() +
  ggtitle("Linear trend") +
  geom_point(
    data = netherlands_data,
    aes(x = year, 
    y = annual_pertussis_cases),
    colour = "red") +
  geom_line(
    data = netherlands_data %>% na.omit, ## note the pipe to remove NA
    aes(x = year, 
        y = annual_pertussis_cases),
    colour = "red",
    linetype = "dashed"
)

```

## How can we apply this to `every` country in the dataset? 
Without doing the above `r pertussis_data_tidy$country %>% as_factor() %>% levels() %>% length()` times

## Split the data by country and apply the model or graphing code to each subset

 - In fact, data for each country is a subset of the full dataset
 - We can subset the original dataframe into seperate dataframes for each country 

## List-columns to track your results and models
```{r}
nested_pertussis <- pertussis_data_tidy %>%
  na.omit() %>%
  dplyr::select(country, year, annual_pertussis_cases) %>%
  group_by(country) %>%
    nest() 
```

## Inspecting the nested dataframes
```{r}
head(nested_pertussis, 2) ## you see the grouping by country
head(nested_pertussis$data[[1]], 2) ## you get the individual country df
```

## Label (name) the idividual elements of the list column 
```{r} 
names(nested_pertussis$data) <- nested_pertussis$country 
head(nested_pertussis$data[1]) 
```
 
## Linear model for each country
First we write a function that creates the linear model for one country
```{r}
country_model_lm <- function(df){
  model <- lm(
    annual_pertussis_cases ~ year, 
    data = df)
  return(model)
}
```

## Iterate the model function over nested `$data` with `purrr::map()`
```{r}
models <- map(
  nested_pertussis$data, country_model_lm
  )
head(models, 2)
```

## Keep it togetheR
 - We have the models now
 - Better to store them together with the data and the group ('country') info
 - By using `dplyr::mutate()` in conjunction with `map()`

[`map()` vs. `lapply()`]https://stackoverflow.com/questions/45101045/why-use-purrrmap-instead-of-lapply)
 
## Create an additional list-column on the basis of an existing one
```{r}
nested_pertussis <- nested_pertussis %>%
  mutate(models_lm = map(data, country_model_lm))
head(nested_pertussis, 2)
```

## Add model summaries as a list-column
```{r}
nested_pertussis <- nested_pertussis %>%
  mutate(models_lm_summary = map(models_lm, summary))
head(nested_pertussis, 2)
```

## Extracting information from a list-column of models; `glance()` & `pluck()`
```{r}
nested_pertussis <- nested_pertussis %>%
  mutate(params_lm =  map(models_lm, broom::glance)) %>%
  mutate(p_value = map(params_lm, pluck, "p.value"))

nested_pertussis$params_lm[[1]]
nested_pertussis$p_value[[1]] %>% round(6)

```

## Adding a list of plots in a column
A function that creates a graph for a single country 
```{r}
plot_line <- function(df, key){
  
  model <- lm(
  annual_pertussis_cases ~ year, 
  data = df %>%
    na.omit()
)
## plot model for NL

plot <- df %>%
  na.omit() %>%
  add_predictions(model) %>%
  ggplot(aes(x = year, 
             y = pred)) +
  geom_line() +
  geom_point(
    data = df,
    aes(x = year, 
    y = annual_pertussis_cases),
    colour = "red") +
  geom_line(
    data = df %>% na.omit, ## note the pipe to remove NA
    aes(x = year, 
        y = annual_pertussis_cases),
    colour = "red",
    linetype = "dashed"
) +
  ggtitle(paste("Linear model for", key %>% as.character()))

  return(plot)
}



```

## Iterate plot function over nested data
```{r}
nested_pertussis <- nested_pertussis %>% 
  mutate(
    plots_lm = map2(data, country, plot_line)
  )
nested_pertussis
```

## Name elements in the list-column
```{r}
names(nested_pertussis$plots_lm) <- nested_pertussis$country
```

## Show a plot
```{r, echo=FALSE, fig.width=6, fig.height=4}
nested_pertussis$plots_lm$Belarus
# nested_pertussis$plots_lm$Belgium
```

## Panel of plots
```{r, echo=FALSE, fig.width=6, fig.height=4}
cowplot::plot_grid(plotlist = 
                     list(nested_pertussis$plots_lm$Netherlands,
                       nested_pertussis$plots_lm$Switzerland)
                                       )

```

## Consider this

Imagine you are writing/using a function to loop over data or models in a list-(column) with `map()` or `lapply`, but it throws an ERROR half way through the list, stopping the loop

How would you solve this? 

The answer is PART II below

## Learn more?

'Managing many models with R' by Hadley Wickham - Lecture
https://www.youtube.com/watch?v=rz3_FDVt9eg

'R for Data Science' by Garret Grolemund & Hadley Wickham
https://r4ds.had.co.nz/
Especially chapters:
21 - https://r4ds.had.co.nz/iteration.html
25 - https://r4ds.had.co.nz/many-models.html

##
```{r, dpi=40, echo=FALSE}
knitr::include_graphics(path = file.path(root,
                                         "images",
                                         "pepper.jpg"))
```



# PART II; Extracting more information from a list column 

**ADVANCED**

For another day...

## Looking at quantative statistical measures for model quality
```{r}
r_squared <- nested_pertussis %>%
  dplyr::filter(country != "Monaco") %>% ## remove Monaco (incomplete data)
  mutate(glance = map(models_lm, broom::glance)) %>% 
  unnest(glance, drop = TRUE) %>%
  select(country, r.squared, AIC, BIC) %>%
  arrange(r.squared)
```

## Plotting r.sqared values
```{r, echo=FALSE, fig.width=8, fig.height=12}
r_squared %>%
  ggplot(aes(x = reorder(as_factor(country), r.squared),
             y = r.squared)) +
  geom_point() +
  coord_flip() +
  ylab("R-squared") +
  xlab("Countries") + 
  geom_point(data = filter(r_squared, r.squared <= 0.25), colour = "red")
```

## Plotting pertussis cases for countries with low r.squared over time
 1. Filter countries with `r.squared` <= 0.25
 2. Put  countries in vector
 3. Plot data

## Step 1
```{r}
low_r_squared <- r_squared %>%
  dplyr::filter(r.squared <= 0.25) %>%
  dplyr::select(country) 
low_r_squared <- low_r_squared$country
```

## Step 2
```{r}
low_r_squared_nested <- nested_pertussis %>%
  dplyr::filter(country %in% low_r_squared) %>%
  select(country, data) %>%
  unnest()
```

## Step 3
```{r, echo=FALSE, fig.width=8, fig.height=12}
low_r_squared_nested %>%
  ggplot(aes(x =year, y = annual_pertussis_cases)) +
  geom_line(aes(group = country, color = country)) +
    theme_bw() +
  theme(legend.position="none") +
  facet_wrap( ~ country, scales = "free") +
 theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

What is happening to the pertussis vaccination grade over the past 8 years? 

## Store ggplot2 objects in a list-column
 1. Create a function that makes the plot
 3. Test function on single dataframe
 2. Apply the function using `mutate()` and `map()` to all dataframes or models
 
```{r}
## isolate one dataframe to test function
df <- nested_pertussis$data[[1]]
plot_country <- function(df){
  
  df %>%
    ggplot(aes(x = year,
           y = annual_pertussis_cases)) +
    geom_line() +
    geom_smooth() +
    ylab("Annual cases")

}

## test function
# plot_country(df = df)
```

## Apply plotting function to nested data
```{r}
nested_pertussis <- nested_pertussis %>%
  mutate(plots_cases_over_time = map(data, 
                                     plot_country))

```

## Add countries as names to new list-column
```{r, fig.width=6, fig.height=4}
names(nested_pertussis$plots_cases_over_time) <- 
  c(nested_pertussis$country)

nested_pertussis$plots_cases_over_time[1]
```

## Pull out "The Netherlands"
```{r, fig.width=6, fig.height=4}
nested_pertussis$plots_cases_over_time$Netherlands # or
# pluck(nested_pertussis$plots_cases_over_time, "Netherlands") + ggtitle("The Netherlands")
```

## Plotting a panel of 4 graphs
```{r, echo=FALSE, fig.width=6, fig.height=4}
grid <- cowplot::plot_grid(
  nested_pertussis$plots_cases_over_time$Armenia, 
  nested_pertussis$plots_cases_over_time$Austria,
  nested_pertussis$plots_cases_over_time$Ukraine,
  nested_pertussis$plots_cases_over_time$Netherlands, 
  labels = c("Armenia", "Austria", "Ukraine", "Netherlands"),
  scale = 0.7, 
  hjust = c(-0.7, -0.8, 
            -0.8, -0.5)) ## aligning the titles is a bit hard
grid
```

## Literature
For some background on the pattern we are seeing

 - https://www.scientificamerican.com/article/why-whooping-cough-vaccines-are-wearing-off/
 - http://outbreaknewstoday.com/pertussis-cases-up-significantly-in-the-eu-netherlands-and-uk-worst-hit-55315/


## Exploring many more models
Let's add a quadratic model in the mix. Assume we want to explore non-linear relationships in this dataset
```{r}
non_linear_model <- function(df, model_params){
  
  nl_model <- lm(
    annual_pertussis_cases ~ poly(year, 
                                  model_params),
                 data = df)
  
  return(nl_model)
}
```

## Creating a safe version of this function
```{r}
safe_non_linear <- purrr::safely(non_linear_model)
## apply test:
df = nested_pertussis$data[[1]]

test_non_linear <-  df %>% non_linear_model(df = .,
                                            model_params = 2)
```

## Test function on one country
```{r}
test_non_linear %>% broom::glance()
```

## Apply model to all countries 
We rerun the steps above to add this new model and new graphs to the nested dataframe

Add new model to the nested table
```{r}
nested_pertussis <- nested_pertussis %>%
  mutate(models_nl_2 = map(data, safe_non_linear, 
                          model_params = 2))
  
nested_pertussis$models_nl_2 <- transpose(nested_pertussis$models_nl_2)
```

## Set names to elements in the list-column
To be able to `pluck()` by name later
```{r}
names(nested_pertussis$models_nl_2$result) <- nested_pertussis$country
```

## Pluck results in new list-column
```{r}
nested_pertussis$models_nl_2$result[[1]] %>% summary

nested_pertussis <- nested_pertussis %>%
  mutate(statistics_nl = pluck(models_nl_2, "result"))

nested_pertussis$statistics_nl[[1]] %>% summary

```

## Tidy the list-column with `{broom}`
```{r}
nested_pertussis <- nested_pertussis %>%
  mutate(parameters_nl = map(statistics_nl, glance))
```


## Looking at quantative statistical measures for model quality
```{r}
r_squared_nl <- nested_pertussis %>%
  select(country, parameters_nl) %>%
  unnest()
```

## Plotting r.sqared values
```{r, echo=FALSE, fig.width=6, fig.height=8}
r_squared %>%
  ggplot(aes(
    x = reorder(as_factor(country), r.squared),
    y = r.squared
  )) +
  geom_point() +
  ylab("Countries") +
  geom_point(data = filter(
    r_squared,
    r.squared <= 0.25
  ), colour = "red") +
  geom_point(
    data = r_squared_nl,
    aes(
      x = reorder(as_factor(country), r.squared),
      y = r.squared
    ), colour = "purple",
    shape = 21, size = 3
  ) +
  theme_bw() +
  xlab("Country") +
  ylab("R.squared") +
  coord_flip() +
  geom_point(data = r_squared_nl %>%
               dplyr::filter(country == "Ireland" |
                             country == "Belgium"), 
             colour = "blue", shape = 22, size = 5)



```

## Let's examine two models for two countries where the non-linear did and did not not improve the R.squared: Ireland (improved) and Belgium (not-improved)  
```{r}
x <- nested_pertussis %>%
  select(country,
         data, 
         models_lm,
         statistics_nl) %>%
  gather(models_lm:statistics_nl, key = "models", value = "model_params") %>%
 print()  
```

## Remove 'empty model'
```{r}
ind <- x$model_params == "NULL"
#ind <- x$data == "NULL"
x <- x[!ind, ]
```

## Add prediction-list column
```{r}
predictions <- x %>%
#  filter(country == "Czech Republic") %>%
  mutate(predictions = map2(data, model_params, add_predictions,
         var = "annual_pertussis_cases")) %>%
  filter(country == "Ireland" | 
         country == "Belgium") %>%
  select(country, data, predictions)

```

## Set names
```{r}
names(predictions$predictions) <- predictions$country
```

## Belgium
```{r, echo=FALSE, fig.width=6, fig.height=4}
nested_pertussis$plots_cases_over_time$Belgium +
  geom_line(data = predictions$predictions[[1]], colour = "red", size = 2) +
  
  geom_line(data = predictions$predictions[[3]], colour = "green", size = 2) 
```  

## Ireland
```{r, echo=FALSE, fig.width=6, fig.height=4}
nested_pertussis$plots_cases_over_time$Ireland+
  geom_line(data = predictions$predictions[[2]], colour = "red", size = 2) +
  
  geom_line(data = predictions$predictions[[4]], colour = "green", size = 2) 
```

## Learn more?
To practice with more examples have a look at
```{r, eval=FALSE}
source(file = file.path(root, "R", "render_help.R"))
help_console(topic = "repurrrsive")
help_console(topic = "purrr")
```

## Disclaimer & Licence

The work presented here may be shared, remixed or adapted as long as the original references and the authors of this document are mentioned in the redistribution: 
LICENCE: CC BY-SA


## Credits

Much of this material has been derived in one way or the other from the work of Hadley Wickham and Garret Grolemund and many others. For a more elaborate reference list see the resources.Rmd file in the project root.

Thanks to Hadley & Garret for writing the book "R for Data Science" http://r4ds.had.co.nz/ and for their work in general to innovate the R world.

Work on integration of Git/Github with R/RStudio is thoroughly and wit-fully documented by Jenny Brian. I also very much appreciate her work on the use of RMarkdown and thanks for pointing me into the direction of using the `rprojroot` package (CSAMA Course 2016). See also:

https://github.com/jennybc/happy-git-with-r & http://stat545.com/block007_first-use-rmarkdown.html 
